\paragraph{Acknowledgments}
This chapter contains contributions by Colin Rothgang.
J\"urgen Sch\"onw\"alder provided comments on the first version.

\section{History and Introduction}\label{sec:sd:crypto:hist}
The definitions in this chapter are based on \cite{PuMaC2016}. 
This section is based partially on \cite{cryptoNetworkSlides}. 
\input{crypto_history}

\section{Fundamental Concepts}

No system is perfectly secure---a brute-force attack is always possible.
Therefore, a central idea of cryptography is that waiting for a polynomial amount of time (in the sense of complexity theory) is feasible but waiting for a super-polynomial amount of time is not.
Then a system is often defined as secure if the best possible attack has super-polynomial complexity.

\begin{definition}[Polynomial and Negligible Functions]
 A function $f:\N\to\R^+$ is called
  \begin{compactitem}
   \item \textbf{polynomial} if $f\in O(p)$ for some polynomial $p$.
   \item \textbf{super-polynomial} if $f\nin O(p)$ for every polynomial $p$ (i.e., if it is not polynomial).
   \item \textbf{negligible} if $f\in O(1/|p(x)|)$ for every polynomial $p:\N\to\R^+$.
  \end{compactitem}
\end{definition}

Super-polynomial functions increase faster than any polynomial increases.
Negligible functions are somewhat dual: they decrease faster than any polynomial increases.
In cryptography, we can think of polynomial/super-polynomial/negligible functions as normal/infeasible/trivial.

\begin{theorem}
We have the following closure properties:
\begin{compactitem}
 \item The sum of super-polynomial/polynomial/negligible functions is super-polynomial/polynomial/negligible again.
 \item The product of a super-polynomial/negligible function with a polynomial function is super-polynomial/negligible again.
 \item A function that is greater/smaller than a super-polynomial/negligible function is super-polynomial/negligible again.
\end{compactitem}
\end{theorem}

\begin{definition}[Polynomial Time]
 An algorithm $A$ is called \textbf{polynomial time} if its worst-case time complexity of $A$ for input of size $n$ is a polynomial function.
\end{definition}

\begin{definition}[Probabilistic Algorithm]
 A \textbf{probabilistic algorithm} is like an algorithm except that it may return different results when called multiple times for the same input.
\end{definition}

In most situations, an algorithm that may return different results for the same input is simply broken and not an algorithm at all.
However, occasionally, probabilistic algorithms are very useful.
Cryptography is an example because we often \emph{need} to make a random choice to prevent an attacker from predicting what we did.
For example, it is often unavoidable for the attacker to know what encryption scheme we use---but that is acceptable if we choose a random key.
Therefore, encryption schemes often consist of a probabilistic key generation algorithm and deterministic encryption/decryption algorithms that take the key as input.
Another application in cryptography is \emph{padding} a message: to prevent leaking the information how long a message is, we can pad every message to a fixed length by adding random data.
In that case, the encryption algorithm is also probabilistic.

Independent of applications in cryptography, probabilistic algorithms are also of great help in situations where (i) a deterministic algorithm has a large complexity but (ii) a probabilistic algorithm that sometimes returns false results is much faster.
If it is easy to test whether a result is correct, we can simply re-run the probabilistic algorithm until it finds a correct result.
That is sometimes used in key generation algorithms, e.g., to find a large prime number.
%Alternatively, if we cannot test if a result is correct, we may be able to run the probabilistic algorithm multiple times in order to increase the probability that a result is correct.

\begin{definition}[Probabilistic Polynomial Time]
 An algorithm is called \textbf{probabilistic polynomial time} (PPT) if it is probabilistic and polynomial.
\end{definition}

One-way-functions (OWF) are functions that are easy to compute but hard to invert:

\newcommand{\Prob}[2]{\mathop{\mathrm{Pr}}_{#1}[#2]}

\begin{notation}
For a function $f:X\to\Bool$, we write $\Prob{x\in X}{f(x)}$ for the probability that $f(x)$ is true when $x$ is chosen uniformly from $X$.
\end{notation}

\begin{definition}[One-Way-Function]
 A polynomial function $f:\Sigma^*\to \Sigma^*$, is a \textbf{one-way-function} if for every probabilistic algorithm $A(n\in\N,y\in \Sigma^*)\to\Sigma^*$, whose complexity is polynomial in $n$, the function
 \[n\tb\mapsto\tb \Prob{x\in\Sigma^n}{f(A(n,f(x)))=f(x)}\]
 is negligible.
\end{definition}
Intuitively, OWFs are super-polynomially hard to invert.
More precisely, any $A$ that attempts to guess (i.e., it is probabilistic) an $x\in\Sigma^n$ such that $y$ and $A(n,y)$ behave the same way under $f$ succeeds with negligible probability.

\begin{example}
It is not known if one-way functions exist.

Some functions that are commonly believed to be OWFs are discrete exponentiation and multiplication.
For example, we do not know if there is a polynomial factoring algorithm.
\end{example}

Pseudo-random-generators (PRG) are functions that can be used to generate numbers that are essentially random:
\begin{definition}[Pseudo-Randomness]
 A polynomial function $R:\Sigma^*\to\Sigma^*$ is a \textbf{pseudo-randomness-generator}
 \begin{compactitem}
   \item $R$ maps input of the same length to output of the same length, i.e., there is an $l:\N\to\N$ such that $R$ maps $\Sigma^n\to\Sigma^{l(n)}$ for every $n$.
   \item for every PPT $A(x\in\Sigma^*)\in\Bool$ the function
  \[n \tb\mapsto \tb \left|\,\Prob{x\in\Sigma^n}{A(R(x))}-\Prob{x\in\Sigma^{l(n)}}{A(x)}\,\right|\]
  is negligible.
 \end{compactitem}
\end{definition}

Intuitively, $R$ is a PRG if it maps $\Sigma^*\to\Sigma^*$ such that no PPT $A$ can tell the difference between $R(x)$ and randomly chosen $x$ with non-negligible probability.
The condition about the length of the output of $R$ is only needed so that we can take the probability on the right-hand side over a finite set.

PRGs and OWFs are intricately linked together:
\begin{theorem}[PRG Theorem]
Given a PRG, we can define an OWF, and vice versa.
\end{theorem}

%  TODO: Insert a proof for at least OWPs using hard core bits
%%This part should probably be moved to the appendix
%\begin{definition}[hard-core bit]
% Let $f$ be a one-way function. Now the function $b:\{0,1\}^*\to\{0,1\}$ is called a \emph{hard-core bit} for $f$ if $b$ is computable in polynomial time and there exists a negligible function $neg(n)$ such that for any $n\in\N$ and any PPT $A$:
% $$\mathrm{Pr}\left[A(f(x),1)=b(x)\right]\leq\frac{1}{2}+neg(n), $$ where $x\in\{0,1\}$ uniformly random. 
%\end{definition}
%
%The next step is to find one-way functions (assuming that there are one-way functions at all) for which we can build hard-core bits. 
%\begin{theorem}
% Let $f$ be a one-way function. Define $g:\{0,1\}^{2n}\to\{0,1\}^*$ as $g(x\circ y)=f(x)\circ y$, where $\left|x\right|=\left|y\right|=n$. Then $g$ is a one-way function with hard-core bit $$b(x,y)=\bigoplus_{i=1}^n x_i\land y_i=\sum_{i=1}^{n}x_iy_i (\mod 2).$$
%\end{theorem}
%If now $f$ was a one-way permutation, we can use the above construction to get an additional pseudo-random bit, so we already have a pseudo-random generator.

\section{Symmetric Encryption}\label{sec:sd:crypto:sym}
\input{symmetric}
% substitution, one-time pad, DES

\section{Asymmetric Encryption}\label{sec:sd:crypto:asym}

\subsection{RSA}

The idea behind RSA is that if $N=p\cdot q$ for large prime numbers $p$ and $q$, it is very difficult to compute $p$ and $q$ from $N$.

\paragraph{Setup}
Choose two large primes $p$ and $q$ (typically of roughly equal size).
Put $N=p\cdot q$.

Now put $n=(p-1)(q-1)$. (Actually, any common multiple of the two numbers is fine.)
Note that $n=\phi(N)$.
Pick $e\in \Z_n$ such that there is a $d\in\Z_n$ with $e\cdot d\Equiv_n 1$.
Such a $d$ exists if $\gcd(e,n)=1$ and is easy to compute (see Thm.~\ref{thm:math:extendedeuclid}).

The keys are defined as follows:
\begin{compactitem}
 \item public information (encryption key): $N$ and $e$
 \item private information (decryption key): $n$, $d$, $p$, and $q$
\end{compactitem}
Among the private information, only $N$ and $d$ are needed later on.
So $n$, $p$, and $q$ can be forgotten.
But they have to remain private---$p$ (or $q$) is enough to compute $n$ and $d$.

Different keys are often compared by their size.
That size is the number of bits in $N$.

\paragraph{Encryption}
Messages are numbers $x\in F_N$.
For example, we can choose the largest $k$ such that $2^k<N$ and use $k$-bit messages.

Encryption and decryption are functions $\Z_N\to \Z_N$ given by
\begin{compactitem}
 \item encryption: $x\mapsto x^e\modop N$
 \item decryption: $x\mapsto x^d\modop N$
\end{compactitem}

These are indeed inverse to each other:

\begin{theorem}
For all $x\in \Z_N$, we have $(x^d)^e\Equiv_N (x^e)^d \Equiv_N x$.
\end{theorem}
\begin{proof}
In general, because $N=p\cdot q$ for prime numbers $p$ and $q$, we have that $x\Equiv_N y$ iff $x\Equiv_p y$ and $x\Equiv_q y$.

So we have to show that $x^{de}\Equiv_p x$.
(We also have to show the same result for $q$, but the proof is the same.)
We distinguish two cases:
\begin{compactitem}
\item $p|x$: Then trivially $x^{de}\Equiv_p x\Equiv_p 0$.
\item Otherwise. Then $p$ and $x$ are coprime.\\
   By construction of $e$ and $d$ and using Thm.~\ref{thm:math:extendedeuclid}, we have $k\in\N$ such that $e\cdot d+k\cdot n=1$.
   Thus, we have to show $x^{de}=x\cdot (x^{p-1})^{k\cdot(q-1)}\Equiv_p x$.
   That follows from $x^{p-1}\Equiv_p 1$ as known from Thm.~\ref{thm:math:fermatlittle}.
\end{compactitem}
\end{proof}

\paragraph{Attacks}
To break RSA, $d$ has to be computed.
There are $3$ natural ways to do that:
\begin{compactitem}
 \item Factor $N$ into $p$ and $q$. Then compute $d$ easily.
 \item Compute $n$ using $n=\phi(N)$ (which may be easier than finding $p$ and $q$). Then compute $d$ easily.
 \item Find $d$ such that $e\cdot d\Equiv_n 1$ (which may be easier than finding $n$).
\end{compactitem}
Currently these are believed to be equally hard.

It is believed that there is no algorithm for factoring $N$ that is polynomial in the number of bits of $N$.
That is not proved.
There are hypothetical machines (e.g., quantum computers) that can factor $N$ polynomially.

Note that checking if $N$ can be factored (without producing the factors) is polynomial, and practical algorithms exist (in particular, the AKS algorithm).
Incidentally, that is important to find the large prime number $p$ and $q$ efficiently.
\medskip

If there is indeed no polynomial algorithm, factoring relies on brute-force attacks that find all prime numbers $k<\sqrt{N}$ and test $k|N$.
Therefore, larger keys are harder than break to smaller ones.
Because of improving hardware, the key size that is considered secure grows over time.

Keys of size $1024$ are considered secure today, but because security is a relative term, keys of size $2048$ are often recommended. 
%It is quetionable that 1024-bit rsa is really secure, even though it has not yet been publicly broken;
%(see for instance https://en.wikipedia.org/wiki/RSA_%28cryptosystem%29#Integer_factorization_and_RSA_problem or https: //www.schneier.com/blog/archives/2007/05/307digit_number.html).
Larger keys are especially important if data is needed to remain secure far into the future, when faster hardware will be available.

\section{Hashing}\label{sec:sd:crypto:hash}

\subsection{MDx}

\subsection{SHA-x}

\section{Authentication}\label{sec:sd:crypto:auth}
   % two-factor authentification

\section{Key Generation and Distribution}

   % security protocol correctness, such as key management protocol correctness,
   % e.g., going back to the work by Burrows, Abadi, and Needham on a logic of authentication (https://doi.org/10.1145/77648.77649)
   % Its an application of logic to a concrete problem space where errors are obviously bad.
